Follow-up Samples

### 2017-04-24 submission/manuscript.tex
re-rendering as requested

\textbf{Results.} There were large changes to the bacterial communities
associated with treatment across the three groups. The communities from
patients with carcinomas changed significantly more than those with
adenoma following treatment (P-value=5.4e-05); however, there was not a
significant difference between those with advanced adenoma and those
with adenoma or carcinoma (P-value \textgreater{} 0.05). Although
treatment brought about large intrapersonal changes, the change in the
abundance of individual OTUs to treatment was not consistent within
diagnosis groups (P-value \textgreater{} 0.05). Because the distribution
of OTUs across patients and diagnosis groups was patchy, we used the
Random Forest machine learning algorithm to identify groups of OTUs that
allowed us to successfully distinguish between pre- and post-treatment
samples for each of the diagnosis groups. However, across the three
models, there was little overlap between the OTUs that were indicative
of treatment. Next, we used a larger cohort that contained individuals
with normal colons and those with adenomas, advanced adenomas, and
carcinomas to determine whether individuals who underwent a treatment
were more likely to have OTUs associated with normal colons. We again
built Random Forest models and measured the change in the positive
probability of having one of the three diagnoses. Although we could
clearly differentiate pre- and post-treatment communities from the three
diagnosis groups, only those patients that initially had carcinomas
experienced a significant decrease in positive probability of having a
carcinoma (P-value \textless{} 0.05). Finally, we tested whether the
type of treatment impacted the microbiota of those diagnosed with
carcinomas and were unable to detect any significant differences in
characteristics of these communities between individuals treated with
surgery alone and those treated with chemotherapy or chemotherapy and
radiation (P-value \textgreater{} 0.05).

@@

Colorectal cancer (CRC) is the third most common cause of cancer deaths
in the United States {[}1,2{]}. Disease mortality has significantly
decreased, thanks to improvements in screening {[}1{]}. Despite this
improvement, there are still approximately 50,000 deaths from the
disease per year {[}2{]}. Current estimates indicate that 20-30\% of
those who undergo treatment will experience recurrence and 30-50\% of
those patients will die {[}3,4{]}. Identification of methods to assess
patients' risk of recurrence is of great importance to reduce mortality
and healthcare costs.

There is growing evidence that the gut microbiota is involved in the
progression of CRC. Mouse-based studies have identified populations of
\emph{Bacteroides fragilis}, \emph{Escherichia coli}, and
\emph{Fusobacterium nucleatum} that alter disease progression
{[}5--9{]}. Furthermore, studies that alter the structure of the
microbiota through the use of antibiotics or inoculation of germ free
mice with human feces has shown that varying community compositions can
result in varied tumor burden {[}10--12{]}. Collectively, these studies
hypothesize that the microbiota is altering the amount of inflammation
in the colon and with it the rate of tumorigenesis {[}13{]}.

Building on this evidence, several human-based studies have identified
unique signatures of colonic lesions {[}14--19{]}. One line of research
has identified community-level differences between those bacteria that
are found on and adjacent to colonic lesions and have supported a role
for \emph{Bacteroides fragilis}, \emph{Escherichia coli}, and
\emph{Fusobacterium nucleatum} in tumorigenesis {[}20--22{]}. Others
have proposed feces-based biomarkers that could be used to diagnose the
presence of colonic adenomas and carcinomas {[}23--25{]}. These studies
have associated \emph{Fusobacterium nucleatum} and other oral pathogens
with lesions (adenoma, advanced adenoma, and carcinoma). They have also
noted that the loss of bacteria generally thought to produce short chain
fatty acids, which can suppress inflammation, is associated with colonic
lesions. Gut bacteria have a role in tumorigenesis and there is promise
that these populations may be useful biomarkers for aiding in the early
detection of disease.

@@

We answered these questions by sequencing the V4 region of 16S rRNA
genes amplified from fecal samples of individuals with adenoma, advanced
adenoma, and carcinomas pre- and post-treatment. We used both classical
community analysis to compare the alpha and beta-diversity of
communities pre- and post treatment. Next, we generated Random Forest
models to identify bacterial populations that were indicative of
treatment for each diagnosis group. Finally, we measured the predictive
probability of diagnosis models to determine whether treatment would
alter the probability of being diagnosed as having a normal colon or the
colon of the original diagnosis group. We found that treatment alters
the composition of the gut microbiota; however, we only observed a shift
from diseased to normal colon for those initially diagnosed with
carcinomas. Understanding how the community responds to these treatments
could be a valuable tool for identifying biomarkers to quantify the risk
of recurrence.

@@

\textbf{\emph{Treatment alters the bacterial community structure of
patients diagnosed with colonic lesions.}} Within our 67-person cohort
we tested whether the microbiota of patients with adenoma (N=22),
advanced adenoma (N=19), or carcinoma (N=26) had any broad differences
between pre- and post-treatment samples {[}Table 1{]}. The structure of
the microbial communities of the pre- and post-treatment samples
differed {[}Figure 1A{]}. We found that the communities obtained pre-
and post-treament among the patients with carcinomas changed
significantly more than those patients with adenoma (P-value = 5.4e-05).
There were no significant differences in the amount of change observed
between the patients with adenoma and advanced adenoma or between the
patients with advanced adenoma and carcinoma (P-value \textgreater{}
0.05). Next, we tested whether there was a consistent direction in the
change in the community structure between the pre and post-treatment
samples for each of the diagnosis groups {[}Figure 1B-D{]}. We only
observed a consistent shift in community structure for the patients with
carcinoma when using a PERMANOVA test (adenoma P-value0.999, advanced
adenoma P-value0.945, and carcinoma P-value0.005). Finally, we measured
the number of observed OTUs, Shannon evenness, and Shannon diversity in
the pre and post-treatment samples and did not observe a significant
change for any of the diagnosis groups (P-value \textgreater{} 0.05)
{[}Table S1{]}.

\textbf{\emph{The effects of treatment are not consistent across
treatment groups.}} We used two approaches to identify those bacterial
populations that change between the two samples for each diagnosis
group. First, we sought to identify individual OTUs that could account
for the change in overall community structure. However, using a paired
Wilcoxon test we were unable to identify any OTUs that were
significantly different in the pre and post-treatment groups {[}Figure
S1{]}. It is likely that high inter-individual variation and the patchy
distribution of OTUs across individuals limited the statistical power of
the test. To overcome these problems we developed Random Forest models
to identify collections of OTUs that would allow us to differentiate
between pre and post-treatment samples from each of the diagnosis
groups. Because of the relatively small number of subjects in each
group, we restricted our models to only incorporate 10 OTUs to limit the
likelihood that that models would overfit the data. Despite this
restriction, the models performed well (adenoma AUC=0.69 - 0.92,
advanced adenoma AUC=0.8 - 1, carcinoma AUC=0.82 - 0.98). Interestingly,
the 10 OTUs that were used for each model had little overlap with each
other {[}Figure 2{]}. These results support the earlier community-wide
analysis where we observed that the treatment had an impact on the
overall community structure; however, the effect of treatment was not
consistent across patients and diagnosis groups.

\textbf{\emph{Post-treatment samples from patients with carcinoma change
towards a microbiota associated with normal}} Next, we determined
whether treatment changed the microbiota in a way that the
post-treatment communities resembled that of patients with normal
colons. To test this, we used an expanded cohort of 423 individuals that
were diagnosed under the same protocol as having normal colons or colons
with adenoma, advanced adenoma, or carcinoma {[}Table 2{]}. We then
constructed Random Forest models to classify the pre and post-treatment
samples as having their original diagnosis or having a normal colon. The
models performed well (adenoma AUC=0.62 - 0.72, advanced adenoma
AUC=0.68 - 0.77, carcinoma AUC=0.84 - 0.9; Figure S2). The OTUs that
were incorporated into the adenoma and advanced adenoma models largely
overlapped and those OTUs that were used to classify the carcinoma
samples were largely distinct from those of the other two models
{[}Figure 3A{]}. Among the OTUs that were shared across the three models
were those populations commonly considered as commensals (e.g.
\emph{Faecalibacterium}, \emph{Lachnospiraceae}, \emph{Bacteroides},
\emph{Dorea}, \emph{Anaerostipes}, and \emph{Roseburia}) {[}Figures
3B{]}. Although many of these OTUs were also included in the model
differentiating between patients with normal colons and those with
carcinoma, this model also included OTUs affiliated with populations
that have previously been associated with carcinoma
(\emph{Fusobacterium}, \emph{Porphyromonas}, \emph{Parvimonas})
{[}Figure S3{]}. Finally, we applied these three models to the pre and
post-treatment samples for each diagnosis group and quantified the
change in the positive probability of the model. A decrease in the
positive probability would indicate that the colon more closely
resembled that of a patient with a normal colon. There was no
significant change in the positive probability for the adenoma or
advanced adenoma groups {[}Figure 4{]}. The positive probability for the
pre and post-treatment samples from patients diagnosed with carcinoma
significantly decreased. In fact, only 6 of the 26 patients (23.08\%)
that were initially diagnosed with a carcinoma had a higher positive
probability after treatment. One of those was re-diagnosed with
carcinoma on the follow up visit. These results indicate that although
there are changes in the microbiota associated with treatment, those
experienced by the patients in the carcinoma group were the only ones
where the change was directed towards what would be found in a normal
colon

\textbf{\emph{Difficult to identify effects of specific treatments on
the change in the microbiota.}} The type of treatment that the patients
received varied across diagnosis groups. Those with adenomas and
advanced adenomas received surgical resection (adenoma, N=4; advanced
adenoma, N=4) or polyp removal during colonoscopy (adenoma, N=18;
advanced adenoma, N=15) and those with carcinomas received surgical
resection (N=12), surgical resection with chemotherapy (N=9), and
surgical resection with chemotherapy and radiation (N=5). We focused on
the patients with carcinoma and pooled those patients that received
chemotherapy with those that received chemotherapy and radiation to
improve our statistical power. We did not observe a significant
difference in the effect of these treatments on the number of observed
OTUs, Shannon diversity, or Shannon evenness (P-value \textgreater{}
0.05). Furthermore, there was not a significant difference in the effect
of the treatments on the amount of change in the community structure (P=
1). Finally, the change in the positive probability was not
significantly different between the two treatment groups (P=1). Due to
the relatively small number of samples in each treatment group, it is
difficult to make a definitive statement regarding the specific type of
treatment on the amount of change in the structure of the microbiota

@@

Understanding the effect of treatment on the microbiota of those
initially diagnosed with carcinomas may have important implications into
understanding disease recurrence. The microbiota changes associated with
treatment among patients initially diagnosed with carcinomas resulted in
community structures that resembled patients with normal colons. This
would suggest that treatment for carcinoma is not only successful for
removing the carcinoma but also at removing bacteria associated with
them. Interestingly, 6 of the 26 patients diagnosed with carcinomas had
an elevated probability of having carcinomas after treatment. This leads
us to hypothesize that these individuals may be at a high risk of
recurrence. Interestingly, the 23.08\% prevalence of increased carcinoma
probability from our study is within the expected rate of recurrence
(20-30\% {[}3,4{]}), which would support our hypothesis . Unfortunately,
because of the design of our study, it was not possible to determine
whether individuals who had an increased positive probability for
carcinoma experienced recurrence. Regardless, it is intriguing to think
of using biomarkers from the microbiota to not only predict the presence
of lesions, but to also assess the risk of developing them.

@@

It was not clear why 6 of the 26 patients with carcinoma had a higher
probability of having carcinomas after treatment. We hypothesized that
these individuals may have had more severe tumors; however, the tumor
severity of these 6 individuals (3 with Stage II and 3 with Stage III)
was similar to the distribution observed among the other 20 patients. We
also hypothesized that we may have sampled these patients later than the
rest and their communities may have reverted to a carcinoma-associated
state; however, there was not a statistically significant difference in
the length of time between sample collection among those whose
probabilities increased or decreased (Wilcoxon Test; P-value=0.56).
Finally, it is possible that these patients may not have responded to
treatment as well as the other 20 patients diagnosed with carcinoma and
so the microbiota may not have been impacted the same way. Again,
further studies looking at the role of the microbiota in recurrence are
needed to understand the dynamics following treatment.

@@

\textbf{\emph{16S rRNA Gene Sequencing.}} Sequencing was completed as
described by Kozich, et al. {[}29{]}. DNA extraction used the 96-well
Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
automated pipetting system (Eppendorf). The V4 variable region was
amplified and the resulting product was split between three sequencing
runs with normal, adenoma, and carcinoma evenly represented on each run.
Each group was randomly assigned to avoid biases based on sample
collection location. The pre- and post-treatment samples were sequenced
on the same run.

\textbf{\emph{Sequence Processing.}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences and has been
previously described {[}29{]}. The general workflow using mothur
included merging paired-end reads into contigs, filtering for low
quality contigs, aligning to the SILVA database {[}reference{]},
screening for chimeras using UCHIME {[}reference{]}, classifying with a
naive Bayesian classifier using the Ribosomal Database Project
(RDP){[}reference-Wang{]}, and clustered into Operational Taxonomic
Units (OTUs) using a 97\% similarity cutoff with an average neighbor
clustering algorithm {[}reference-Westcott\&SchlossAEM{]}. The number of
sequences for each sample was rarefied to 10523 to minimize the impacts
of uneven sampling.

\textbf{\emph{Model Building.}} The Random Forest {[}30{]} algorithm was
used to create the model used to create the three models used. The
adenoma model classified normal versus adenoma, advanced adenoma was
normal versus advanced adenoma, and carcinoma was normal versus
carcinoma. The toal number of individuals in this data set was 423
individuals {[}Table 1{]}. There were a total of 239 individuals in the
adenoma model, 262 individuals in the advanced adenoma model, and 266
indivdiuals in the carcinoma model {[}Table 1{]}. Each model was then
applied to our 67-person cohort testing prediction of adenoma intial
(adenoma n = 22) versus adenoma follow up (adenoma n = 0), advanced
adenoma initial (advanced adenoma n = 19) versus advanced adenoma follow
up (advanced adenoma n = 0), carcinoma initial (carinoma n = 26) versus
carcinoma follow up (carcinoma n = 1).

@@

\textbf{\emph{Statistical Analysis.}} The R software package (v3.3.2)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}31{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall OTU
differences between pre- and post-treatment samples utilized a paired
Wilcoxon ranked sum test. Where multiple comparison testing was
appropriate, a Benjamini-Hochberg (BH) correction was applied {[}32{]}
and a corrected P-value of less than 0.05 was considered significant.
Unless otherwise stated the P-values reported are those that were BH
corrected. Model rank importance was determined by obtaining the median
MDA from the 100, 20 repeated 10-fold cross validation and then ranking
from largest to smallest MDA.

@@

\textbf{Figure 1: General Differences between Adenoma, Advanced Adenoma,
and Carcinoma Groups After Treatment.} A) A significant difference was
found between the adenoma and carcinoma group for thetayc (P-value =
NULL). Solid black points represent the median value for each diagnosis
group. B) NMDS of the pre- and post-treatment samples for the adenoma
group. C) NMDS of the pre- and post-treatment samples for the advanced
adenoma group. D) NMDS of the pre- and post-treatment samples for the
carcinoma group.

@@

\textbf{Table 1: Demographic Data of Pre and Post Treatment Cohort}

@@

\hypertarget{ref-breiman_random_2001}{}
30. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
{[}cited 2013 Feb 7{]};45:5--32. Available from:
\href{http://link.springer.com/article/10.1023/A\%3A1010933404324\%20http://link.springer.com/article/10.1023\%2FA\%3A1010933404324?LI=true}{http://link.springer.com/article/10.1023/A\%3A1010933404324 http://link.springer.com/article/10.1023\%2FA\%3A1010933404324?LI=true}

\hypertarget{ref-anderson_permanova_2013}{}
31. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
the face of heterogeneous dispersions: What null hypothesis are you
testing? Ecological Monographs {[}Internet{]}. 2013 {[}cited 2017 Jan
5{]};83:557--74. Available from:
\url{http://doi.wiley.com/10.1890/12-2010.1}

\hypertarget{ref-benjamini_controlling_1995}{}
32. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
practical and powerful approach to multiple testing. Journal of the
Royal Statistical Society. Series B (Methodological). 1995;57:289--300.

>>>>

\textbf{Results.} There were large changes to the bacterial communities
associated with treatment across the three groups. The communities from
patients with carcinomas changed significantly more than those with
adenoma following treatment (P-value \textless{} 0.001); however, there
was not a significant difference between those with advanced adenoma and
those with adenoma or carcinoma (P-value \textgreater{} 0.05). Although
treatment brought about large intrapersonal changes, the change in the
abundance of individual OTUs to treatment was not consistent within
diagnosis groups (P-value \textgreater{} 0.05). Because the distribution
of OTUs across patients and diagnosis groups was patchy, we used the
Random Forest machine learning algorithm to identify groups of OTUs that
allowed us to successfully distinguish between pre and post-treatment
samples for each of the diagnosis groups. Although the three models
successfully differentiated between the pre and post-treatment samples,
there was little overlap between the OTUs that were indicative of
treatment. Next, we used a larger cohort that contained individuals with
normal colons and those with adenomas, advanced adenomas, and carcinomas
to determine whether individuals who underwent a treatment were more
likely to have OTUs associated with normal colons. We again built Random
Forest models and measured the change in the positive probability of
having one of the three diagnoses. Only those patients that initially
had carcinomas experienced a significant decrease in positive
probability of having a lesion (P-value \textless{} 0.05). Finally, we
tested whether the type of treatment impacted the microbiota of those
diagnosed with carcinomas and were unable to detect any significant
differences in characteristics of these communities between individuals
treated with surgery alone and those treated with chemotherapy or
chemotherapy and radiation (P-value \textgreater{} 0.05).

@@

Colorectal cancer (CRC) is the third most common cause of cancer deaths
in the United States {[}1,2{]}. Disease mortality has significantly
decreased, thanks to improvements in screening {[}1{]}. Despite this
improvement, there are still approximately 50,000 CRC-related deaths per
year {[}2{]}. Current estimates indicate that 20-30\% of those who
undergo treatment will experience recurrence and 30-50\% of those
patients will die {[}3,4{]}. Identification of methods to assess
patients' risk of recurrence is of great importance to reduce mortality
and healthcare costs.

There is growing evidence that the gut microbiota is involved in the
progression of CRC. Mouse-based studies have identified populations of
\emph{Bacteroides fragilis}, \emph{Escherichia coli}, and
\emph{Fusobacterium nucleatum} that alter disease progression
{[}5--9{]}. Furthermore, studies that alter the structure of the
microbiota through the use of antibiotics or inoculation of germ free
mice with human feces has shown that varying community compositions can
result in varied tumor burden {[}10--12{]}. Collectively, these studies
support the hypothesis that the microbiota can alter the amount of
inflammation in the colon and with it the rate of tumorigenesis
{[}13{]}.

Building upon this evidence, several human-based studies have identified
unique signatures of colonic lesions {[}14--19{]}. One line of research
has identified community-level differences between those bacteria that
are found on and adjacent to colonic lesions and have supported a role
for \emph{Bacteroides fragilis}, \emph{Escherichia coli}, and
\emph{Fusobacterium nucleatum} in tumorigenesis {[}20--22{]}. Others
have proposed feces-based biomarkers that could be used to diagnose the
presence of colonic adenomas and carcinomas {[}23--25{]}. These studies
have associated \emph{Fusobacterium nucleatum} and other oral pathogens
with colonic lesions (adenoma, advanced adenoma, and carcinoma). They
have also noted that the loss of bacteria generally thought to produce
short chain fatty acids, which can suppress inflammation, is associated
with colonic lesions. Gut bacteria have a role in tumorigenesis and
there is promise that these populations may be useful biomarkers for
aiding in the early detection of disease.

@@

We answered these questions by sequencing the V4 region of 16S rRNA
genes amplified from fecal samples of individuals with adenoma, advanced
adenoma, and carcinomas pre and post-treatment. We used classical
community analysis to compare the alpha and beta-diversity of
communities pre and post treatment. Next, we generated Random Forest
models to identify bacterial populations that were indicative of
treatment for each diagnosis group. Finally, we measured the predictive
probability of diagnosis models to determine whether treatment would
alter the probability of being diagnosed as having a normal colon or the
colon of the original diagnosis group. We found that treatment alters
the composition of the gut microbiota; however, we only observed a shift
from diseased to normal colon for those initially diagnosed with
carcinomas. Understanding how the community responds to these treatments
could be a valuable tool for identifying biomarkers to quantify the risk
of recurrence.

@@

\textbf{\emph{Treatment alters the bacterial community structure of
patients diagnosed with colonic lesions.}} Within our 67-person cohort
we tested whether the microbiota of patients with adenoma (N=22),
advanced adenoma (N=19), or carcinoma (N=26) had any broad differences
between pre and post-treatment samples {[}Table 1{]}. The structure of
the microbial communities of the pre and post-treatment samples differed
{[}Figure 1A{]}. We found that the communities obtained pre and
post-treament among the patients with carcinomas changed significantly
more than those patients with adenoma (P-value = 5.4e-05). There were no
significant differences in the amount of change observed between the
patients with adenoma and advanced adenoma or between the patients with
advanced adenoma and carcinoma (P-value \textgreater{} 0.05). Next, we
tested whether there was a consistent direction in the change in the
community structure between the pre and post-treatment samples for each
of the diagnosis groups {[}Figure 1B-D{]}. We only observed a consistent
shift in community structure for the patients with carcinoma when using
a PERMANOVA test (adenoma P-value=0.999, advanced adenoma P-value=0.945,
and carcinoma P-value=0.005). Finally, we measured the number of
observed OTUs, Shannon evenness, and Shannon diversity in the pre and
post-treatment samples and did not observe a significant change for any
of the diagnosis groups (P-value \textgreater{} 0.05) {[}Table S1{]}.

\textbf{\emph{The effects of treatment are not consistent across
treatment groups.}} We used two approaches to identify those bacterial
populations that change between the two samples for each diagnosis
group. First, we sought to identify individual OTUs that could account
for the change in overall community structure. However, using a paired
Wilcoxon test we were unable to identify any OTUs that were
significantly different in the pre and post-treatment groups {[}Figure
S1{]}. It is likely that high inter-individual variation and the patchy
distribution of OTUs across individuals limited the statistical power of
the test. To overcome these problems we developed Random Forest models
to identify collections of OTUs that would allow us to differentiate
between pre and post-treatment samples from each of the diagnosis
groups. Because of the relatively small number of subjects in each
group, we restricted our models to only incorporate 10 OTUs to limit the
likelihood that the models would overfit the data. Despite this
restriction, the models performed well (adenoma AUC=0.69 - 0.92,
advanced adenoma AUC=0.80 - 1.00, carcinoma AUC=0.82 - 0.98).
Interestingly, the 10 OTUs that were used for each model had little
overlap with each other {[}Figure 2{]}. These results support the
earlier community-wide analysis where we observed that the treatment had
an impact on the overall community structure; however, the effect of
treatment was not consistent across patients and diagnosis groups.

\textbf{\emph{Post-treatment samples from patients with carcinoma change
towards a microbiota associated with normal}} Next, we determined
whether treatment changed the microbiota in a way that the
post-treatment communities resembled that of patients with normal
colons. To test this, we used an expanded cohort of 423 individuals that
were diagnosed under the same protocol as having normal colons or colons
with adenoma, advanced adenoma, or carcinoma {[}Table 2{]}. We then
constructed Random Forest models to classify the pre and post-treatment
samples as having their original diagnosis or having a normal colon. The
models performed well (adenoma AUC=0.62 - 0.72, advanced adenoma
AUC=0.68 - 0.77, carcinoma AUC=0.84 - 0.90; Figure S2). The OTUs that
were incorporated into the adenoma and advanced adenoma models largely
overlapped and those OTUs that were used to classify the carcinoma
samples were largely distinct from those of the other two models
{[}Figure 3A{]}. Among the OTUs that were shared across the three models
were those populations commonly considered as commensals (e.g.
\emph{Faecalibacterium}, \emph{Lachnospiraceae}, \emph{Bacteroides},
\emph{Dorea}, \emph{Anaerostipes}, and \emph{Roseburia}) {[}Figures
3B{]}. Although many of these OTUs were also included in the model
differentiating between patients with normal colons and those with
carcinoma, this model also included OTUs affiliated with populations
that have previously been associated with carcinoma
(\emph{Fusobacterium}, \emph{Porphyromonas}, \emph{Parvimonas})
{[}Figure S3{]}. Finally, we applied these three models to the pre and
post-treatment samples for each diagnosis group and quantified the
change in the positive probability of the model. A decrease in the
positive probability would indicate that the colon more closely
resembled that of a patient with a normal colon. There was no
significant change in the positive probability for the adenoma or
advanced adenoma groups {[}Figure 4{]}. The positive probability for the
pre and post-treatment samples from patients diagnosed with carcinoma
significantly decreased with treatment. In fact, 6 of the 26 patients
(23.08\%) that were initially diagnosed with a carcinoma had a higher
positive probability after treatment. One of those was re-diagnosed with
carcinoma on the follow up visit. These results indicate that although
there are changes in the microbiota associated with treatment, those
experienced by the patients in the carcinoma group were the only ones
where the change was directed towards what would be found in a normal
colon.

\textbf{\emph{Difficult to identify effects of specific treatments on
the change in the microbiota.}} The type of treatment that the patients
received varied across diagnosis groups. Those with adenomas and
advanced adenomas received surgical resection (adenoma, N=4; advanced
adenoma, N=4) or polyp removal during colonoscopy (adenoma, N=18;
advanced adenoma, N=15) and those with carcinomas received surgical
resection (N=12), surgical resection with chemotherapy (N=9), and
surgical resection with chemotherapy and radiation (N=5). We focused on
the patients with carcinoma and pooled those patients that received
chemotherapy with those that received chemotherapy and radiation to
improve our statistical power. We did not observe a significant
difference in the effect of these treatments on the number of observed
OTUs, Shannon diversity, or Shannon evenness (P-value \textgreater{}
0.05). Furthermore, there was not a significant difference in the effect
of the treatments on the amount of change in the community structure (P=
1). Finally, the change in the positive probability was not
significantly different between the two treatment groups (P=1). Due to
the relatively small number of samples in each treatment group, it was
difficult to make a definitive statement regarding the specific type of
treatment on the amount of change in the structure of the microbiota.

@@

Understanding the effect of treatment on the microbiota of those
initially diagnosed with carcinomas may have important implications into
understanding disease recurrence. The microbiota changes associated with
treatment among patients initially diagnosed with carcinomas resulted in
community structures that resembled patients with normal colons. This
would suggest that treatment for carcinoma is not only successful for
removing the carcinoma but also at removing bacteria associated with
them. Interestingly, 6 of the 26 patients diagnosed with carcinomas had
an elevated probability of having carcinomas after treatment. This leads
us to hypothesize that these individuals may be at a high risk of
recurrence. Interestingly, the 23.08\% prevalence of increased carcinoma
probability from our study is within the expected rate of recurrence
(20-30\% {[}3,4{]}). Unfortunately, because of the design of our study,
it was not possible to determine whether individuals who had an
increased positive probability for carcinoma experienced recurrence.
Regardless, it is intriguing that it may be possible to use
microbiome-based biomarkers to not only predict the presence of lesions,
but to also assess the risk of developing them.

@@

It was not clear why 6 of the 26 patients with carcinoma had a higher
probability of having carcinomas after treatment. We hypothesized that
these individuals may have had more severe tumors; however, the tumor
severity of these 6 individuals (3 with Stage II and 3 with Stage III)
was similar to the distribution observed among the other 20 patients. We
also hypothesized that we may have sampled these patients later than the
rest and their communities may have reverted to a carcinoma-associated
state; however, there was not a statistically significant difference in
the length of time between sample collection among those whose
probabilities increased or decreased (Wilcoxon Test; P-value = 0.56).
Finally, it is possible that these patients may not have responded to
treatment as well as the other 20 patients diagnosed with carcinoma and
so the microbiota may not have been impacted the same way. Again,
further studies looking at the role of the microbiota in recurrence are
needed to understand the dynamics following treatment.

@@

\textbf{\emph{16S rRNA Gene Sequencing.}} Sequencing was completed as
described by Kozich, et al. {[}29{]}. DNA extraction used the 96-well
Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
automated pipetting system (Eppendorf). The V4 variable region was
amplified and the resulting product was split between three sequencing
runs with normal, adenoma, and carcinoma evenly represented on each run.
Each group was randomly assigned to avoid biases based on sample
collection location. The pre and post-treatment samples were sequenced
on the same run.

\textbf{\emph{Sequence Processing.}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences and has been
previously described {[}29{]}. The general workflow using mothur
included merging paired-end reads into contigs, filtering for low
quality contigs, aligning to the SILVA database {[}30{]}, screening for
chimeras using UCHIME {[}31{]}, classifying with a naive Bayesian
classifier using the Ribosomal Database Project (RDP){[}32{]}, and
clustered into Operational Taxonomic Units (OTUs) using a 97\%
similarity cutoff with an average neighbor clustering algorithm
{[}33{]}. The number of sequences for each sample was rarefied to 10523
to minimize the impacts of uneven sampling.

\textbf{\emph{Model Building.}} The Random Forest {[}34{]} algorithm was
used to create the model used to create the three models used. The
adenoma model classified normal versus adenoma, advanced adenoma was
normal versus advanced adenoma, and carcinoma was normal versus
carcinoma. The toal number of individuals in this data set was 423
individuals {[}Table 1{]}. There were a total of 239 individuals in the
adenoma model, 262 individuals in the advanced adenoma model, and 266
indivdiuals in the carcinoma model {[}Table 1{]}. Each model was then
applied to our 67-person cohort testing prediction of adenoma intial
(adenoma n = 22) versus adenoma follow up (adenoma n = 0), advanced
adenoma initial (advanced adenoma n = 19) versus advanced adenoma follow
up (advanced adenoma n = 0), carcinoma initial (carinoma n = 26) versus
carcinoma follow up (carcinoma n = 1).

@@

\textbf{\emph{Statistical Analysis.}} The R software package (v3.3.2)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}35{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall OTU
differences between pre and post-treatment samples utilized a paired
Wilcoxon ranked sum test. Where multiple comparison testing was
appropriate, a Benjamini-Hochberg (BH) correction was applied {[}36{]}
and a corrected P-value of less than 0.05 was considered significant.
Unless otherwise stated the P-values reported are those that were BH
corrected. Model rank importance was determined by obtaining the median
MDA from the 100, 20 repeated 10-fold cross validation and then ranking
from largest to smallest MDA.

@@

\textbf{Figure 1: General Differences between Adenoma, Advanced Adenoma,
and Carcinoma Groups After Treatment.} A) A significant difference was
found between the adenoma and carcinoma group for thetayc (P-value =
NULL). Solid black points represent the median value for each diagnosis
group. B) NMDS of the pre and post-treatment samples for the adenoma
group. C) NMDS of the pre and post-treatment samples for the advanced
adenoma group. D) NMDS of the pre and post-treatment samples for the
carcinoma group.

@@

\textbf{Table 1: Demographic Data of pre and Post Treatment Cohort}

@@

\hypertarget{ref-pruesse_silva_2007}{}
30. Pruesse E, Quast C, Knittel K, Fuchs BM, Ludwig W, Peplies J, et al.
SILVA: A comprehensive online resource for quality checked and aligned
ribosomal RNA sequence data compatible with ARB. Nucleic Acids Research.
2007;35:7188--96.

\hypertarget{ref-edgar_uchime_2011}{}
31. Edgar RC, Haas BJ, Clemente JC, Quince C, Knight R. UCHIME improves
sensitivity and speed of chimera detection. Bioinformatics (Oxford,
England). 2011;27:2194--200.

\hypertarget{ref-wang_naive_2007}{}
32. Wang Q, Garrity GM, Tiedje JM, Cole JR. Naive Bayesian classifier
for rapid assignment of rRNA sequences into the new bacterial taxonomy.
Applied and Environmental Microbiology. 2007;73:5261--7.

\hypertarget{ref-schloss_assessing_2011}{}
33. Schloss PD, Westcott SL. Assessing and improving methods used in
operational taxonomic unit-based approaches for 16S rRNA gene sequence
analysis. Applied and Environmental Microbiology. 2011;77:3219--26.

\hypertarget{ref-breiman_random_2001}{}
34. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
{[}cited 2013 Feb 7{]};45:5--32. Available from:
\href{http://link.springer.com/article/10.1023/A\%3A1010933404324\%20http://link.springer.com/article/10.1023\%2FA\%3A1010933404324?LI=true}{http://link.springer.com/article/10.1023/A\%3A1010933404324 http://link.springer.com/article/10.1023\%2FA\%3A1010933404324?LI=true}

\hypertarget{ref-anderson_permanova_2013}{}
35. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
the face of heterogeneous dispersions: What null hypothesis are you
testing? Ecological Monographs {[}Internet{]}. 2013 {[}cited 2017 Jan
5{]};83:557--74. Available from:
\url{http://doi.wiley.com/10.1890/12-2010.1}

\hypertarget{ref-benjamini_controlling_1995}{}
36. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
practical and powerful approach to multiple testing. Journal of the
Royal Statistical Society. Series B (Methodological). 1995;57:289--300.
